%% nilHMM methods paper -- optimization supplement.
%% The C++/Rcpp reimplementation of the RTIGER and nNIL cores: a prerequisite for
%% simulation-scale calibration, established correctness-first (equivalence to the
%% Julia RTIGER), then throughput/memory. Structure mirrors the fork report
%% (faustovrz/RTIGER docs/optimization.md) reframed as C++/Rcpp vs Julia.
%% Compiles standalone or is \input into main.tex's \section*{Supplement}.
%% SCAFFOLD: benchmark numbers are placeholders (see \todo marks); fill from
%% analysis/rtiger-optimization-benchmark.qmd and analysis/nnil-reproduction-benchmark.qmd.
\documentclass[11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{array}
\usepackage[round,authoryear]{natbib}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
\usepackage{xcolor}
\providecommand{\todo}[1]{{\footnotesize\textcolor{red}{[TODO: #1]}}}
\newcommand{\code}[1]{\texttt{#1}}
\setlength{\emergencystretch}{3em}

\title{Supplemental Text: the C++/Rcpp reimplementation\\[2pt]
\large Simulation-calibrated HMM ancestry calling for near-isogenic lines\\
across sequencing modalities (nilHMM)}
\author{Rodr\'iguez-Zapata \textit{et al.}}
\date{}

\begin{document}
\maketitle

%======================================================================
\section{The C++/Rcpp reimplementation of the RTIGER and nNIL cores}
\label{supp:optimization}
%======================================================================

The named callers of the main text run on a single engine reimplemented in
C++/Rcpp (\code{nilhmm/src/}: \code{rtiger.cpp}, \code{viterbi\_batch.cpp},
\code{emission\_count.cpp}, \code{lbimpute.cpp}, \code{fsfhap.cpp}, and the shared
forward/backward and segment routines). This reimplementation is a prerequisite,
not a contribution: it removes the Julia runtime that upstream RTIGER
\citep{rtiger_2023} requires through \code{JuliaCall}, and it makes a single fit
cheap enough to loop over the calibration sweeps and coverage sweeps of the main
text at $10^4$ to $10^5$ markers on a laptop. We establish it correctness-first,
then report speed and memory.

The method here (a shared-panel marker sweep, a Viterbi-identical equivalence
oracle, and a peak-RSS sweep) follows the earlier optimized-Julia benchmark of the
performance fork \code{faustovrz/RTIGER@optimize-julia-core}, whose numbers
(an $O(N^2)\!\to\!O(N)$ reduction in markers, and peak memory made flat in sample
count) are the reference point against which the C++ port is measured.

\paragraph{Data.} Real \emph{Arabidopsis} Col-0 $\times$ Ler allele counts
(BioProject PRJNA720439 / SRA study SRP313911), the 109{,}703-locus panel of loci
covered in all three of the BN/Z/AU samples, decimated by odd index to five sizes
(6{,}857; 13{,}713; 27{,}426; 54{,}852; 109{,}703 markers per sample). The nNIL
reproduction (Sec.~\ref{supp:nnil}) uses the Zhong \emph{et al.} nested-NIL panel.

%----------------------------------------------------------------------
\subsection{Equivalence to Julia RTIGER (correctness first)}
%----------------------------------------------------------------------
Both cores were run to full convergence from identical deterministic
initialization on the shared panel at all five sizes, and their outputs compared
position by position.

\begin{table}[htbp]
\centering
\small
\begin{tabular}{@{}r c c c c@{}}
\toprule
markers / sample & EM iters (C++ / Julia) & Viterbi match & mismatches & param max-diff \\
\midrule
6{,}857   & \todo{} & \todo{} & \todo{0?} & \todo{} \\
13{,}713  & \todo{} & \todo{} & \todo{0?} & \todo{} \\
27{,}426  & \todo{} & \todo{} & \todo{0?} & \todo{} \\
54{,}852  & \todo{} & \todo{} & \todo{0?} & \todo{} \\
109{,}703 & \todo{} & \todo{} & \todo{0?} & \todo{} \\
\bottomrule
\end{tabular}
\caption{nilHMM C++/Rcpp vs Julia RTIGER on the shared \emph{Arabidopsis} panel:
identical EM iteration count, Viterbi-identical ancestry paths (0 mismatches
target), parameters agreeing to float-summation tolerance. Compute:
\code{analysis/rtiger-optimization-benchmark.qmd}.}
\label{tab:cpp-equiv}
\end{table}

%----------------------------------------------------------------------
\subsection{Throughput and memory}
%----------------------------------------------------------------------
Per-iteration and total wall time against markers per sample (fixed iteration
count, native build), and peak RSS as the sample count grows at fixed
markers/sample. \todo{insert marker-scaling figure and the empirical complexity
exponent (log-log slope) for each core; insert peak-RSS table and state whether
memory is flat in sample count.}

%======================================================================
\section{nNIL reproduction and the calibration foil}
\label{supp:nnil}
%======================================================================
nilHMM's \code{nNIL} caller (categorical genotype-confusion emission, geometric
duration) reimplements Holland's original nested-NIL HMM \citep{zhong2025}. Two
results: (i) run on the Zhong \emph{et al.} panel it reproduces the published
introgression calls \todo{marker concordance / donor-segment Jaccard}; (ii) on the
24-line SNP-chip-truth subset, the operating point Zhong \emph{et al.} selected by
chip optimization and the \code{rrate} selected by \code{simcross} calibration (no
chip) coincide \todo{show the two operating points and their segmentation
agreement}. Result (ii) is the concrete population-scale evidence for the main
text's claim that simulation substitutes for scarce high-coverage truth. Compute:
\code{analysis/nnil-reproduction-benchmark.qmd}.

\end{document}
